---
title: "Analyse DNA-seq data using CloneDetective"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{analyse_dnaseq_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette focuses on using the package to analyse DNA-seq data.
DNA-seq data refers to lineage tracing DNA barcoding data which exclusively sequences 
clone barcode reads using Next Generation Sequencing.

We will go through how to analyse clone barcode distributions from DNA-seq data which 
can inform the design of future scRNAseq experiments.

# Setup

In addition to CloneDetective we’ll load the following packages:

* **data.table**, to read in CSV file and manipulate them.
* **ggplot2**, in case we want to customise the ggplot objects created by CloneDetective.
* **scales**, in case we want to customise the scale of the ggplot objects created by CloneDetective.

```{r setup}
library(CloneDetective)
library(data.table)
library(ggplot2)
library(scales)
```

# Load the DNA-seq data

Before we delve into investigating what our DNA-seq data looks like, we have to load it in first.

Assuming you have used [NextClone](https://github.com/phipsonlab/NextClone) to extract 
the clone barcodes from your raw FASTQ files,
you will now have a CSV file where each row corresponds to a unique clone barcode 
and its frequency in the sample. 

If you have used other pipeline to extract (which you are more than welcomed to),
you will have to wrangle the output such that you get the aforementioned CSV file.

For the purpose of this vignette, I have randomly sampled 4000 clone barcodes
from the 8k and 10k DNA-seq data we used in our paper.

Let's load it up using `data.table` and take a peek inside using the `head` function.
The data is included in the package in `extdata` folder which we can located using
`system.file` function.

```{r}
clones_nxclone <- fread(
  system.file("extdata", "dnaseq_clone_barcode_counts_subsampled.csv", 
              package = "CloneDetective")
)


head(clones_nxclone)
```

Let's quickly go through what the columns are:

* **sample_name**: the sample the clone barcode came from. We have 2 samples for this
dataset, 8k and 10k. 
* **known_barcode**: the clone barcodes in our data.
* **read_count**: the number of reads detected for each clone. Since we only sequence
the clone barcode, you can treat this as the number of cells associated with a given
clone barcode.

# Analyse clone barcode distributions

CloneDetective offers functions to count the number of clone barcodes in the dataset 
and to filter out clones with low abundance based on a predefined cell count threshold. 

To do this, we have to first specify the filtering thresholds, i.e. the numeric vector listing 
the minimum cell counts required for a clone to be included in the count. 
For this example, we will set the thresholds to 1, 20, 200, 1000 to see how many clones
we get with at least 1, 20, 200, 1000 cells.

```{r}
thresholds <- c(1, 20, 200, 1000)
```

Then we will count how many clones with 1, 20, 200, 1000 cells.

```{r}
n_barcodes_nxclone <- count_retained_clones(
    count_data = clones_nxclone,
    thresholds = thresholds,
    grouping_col = "sample_name",
    count_column = "read_count"
)
```

The `grouping_col` parameter essentially tells the function to repeat the counting procedure for 
each unique value in the **sample_name** column of our clone data (`clones_nxclone`).
The `count_column` tells the function which column contains the number of cells detected
for each clone.

Let's take a peak at the result.

```{r}
n_barcodes_nxclone
```

Each row is a sample, and each column tells you how many clones we get with at least
1, 20, 200, 1000 cells.
Remember, we are using sub-sampled DNA-seq data, that is why the `at_least_1_cells` column
for both samples have exactly 4,000 clones.

This information is useful to see how many reasonably sized clones we get in our DNA-seq data.
Often, it is not even worthwhile keeping clones that have too little cells.
To remove them, we can use the `remove_clones_below_threshold` function.

```{r}
clones_nxclone_filtered <- remove_clones_below_threshold(
    count_data = clones_nxclone,
    threshold = 20,
    count_column = "read_count"
)
```

Let's see how many clones were removed using the `table` function which is a neat
little function that count the number of occurences in a vector.

```{r}
table(clones_nxclone_filtered$sample_name)
```

While these are all super useful, it is even better if we can look at the proportion of cells 
in each clone (against the total number of cells in the sample), and illustrate this 
as an ordered abundance plot. 
Such plots are insightful for determining whether clones proliferated uniformly or 
if some clones were more dominant, potentially outcompeting and restricting the expansion of others.

We can draw the plot using `draw_ordered_abundance_plot`, but before we do that,
we need to convert our the cell count to proportion.
We can do this using `convert_count_to_proportion` function.
Let's run all of that and add a bit of customisation to our plot to make it look pretty.

```{r}
clones_nxclone_filtered <- convert_count_to_proportion(
  count_data = clones_nxclone_filtered,
  grouping_col = "sample_name",
  count_column = "read_count"
)


plt <- draw_ordered_abundance_plot(
  count_data = clones_nxclone_filtered, 
  facet_column = "sample_name",
  y_axis_column = "read_proportion"
)
plt <- plt +
  geom_point(size=0.5, colour='red') +
  labs(
    title = "Proportion of cells assigned to each barcode",
    subtitle = "Barcode IDs are numerically assigned in order of read proportion",
    x = "Numerical barcode ID",
    y = "Proportion of cells per sample"
  )
plt
```

From this plot, you can see that the 10k sample looks even, which means the clones
are expanding evenly.
However, for the 8k sample, we have one clone (one outlier dot at around 0.015 mark
at the y-axis) which seems to have expanded much more than others.
The proportion of cells belonging to this clone is much much more than others.

For the future, it may be best to not use the 8k sample because of this
dominant clone restricting the space available for others to grow.

# Using DNA-seq data to plan future scRNAseq experiment

We can use the calculated cell proportions above to forecast clonal abundance for 
future scRNAseq experiments. 
This is helpful for estimating the number of cells to sequence for a given clonal 
library in order to maximise the chance of obtaining a reasonable representation of clones. 
For example, say we plan to sequence 10,000 or 20,000 cells.
Can we reasonably estimate how many cells will be obtained per clone? Yes we can.

First thing first though, we need to do the estimation.
We can do this using the `project_clones` function.

```{r}
clones_nxclone_proportion <- project_clones(
  count_data = clones_nxclone,
  grouping_col = "sample_name",
  count_column = "read_count",
  project_amnt = c(10000, 20000)
)
```

You may have noticed by now that many functions in the package use the same parameter names.
They indeed do, and we intentionally do this so to keep things simple.
The only new parameter here is `project_amnt` which if you look at the values, you
may have guessed what they represent, the number of cells to project our cell proportions to.

Let's take quick peek into this.

```{r}
head(clones_nxclone_proportion)
```

We have 2 extra columns now (the **read_count_proportion** column was computed using
`convert_count_to_proportion` in the previous section).
Each column represents the estimated number of cells if you sequence 10,000 or 20,000 cells.

Note, for this estimation, we assume all the cells will get a clone barcode.
Chances are this won't happen, and you will probably get 80% to 90% of cells getting
clone barcode.
We can add this information in as the `confidence_threshold` parameter, and redo the estimation.
Note that `confidence_threshold` takes value between 0 to 1, so you will need to convert
the percentage (if 80% then it is 80/100 = 0.8).

```{r}
clones_nxclone_proportion <- project_clones(
  count_data = clones_nxclone,
  grouping_col = "sample_name",
  count_column = "read_count",
  project_amnt = c(10000, 20000),
  confidence_threshold = 0.8
)

head(clones_nxclone_proportion)
```

We can no visualise this as ordered abundance plot to see what the clone abundance look like.

```{r}
plt <- draw_ordered_abundance_plot(
  count_data = clones_nxclone_proportion,
  y_axis_column = 'projected_20000_confidence_0_8',
  facet_column = 'sample_name'
) + 
  geom_point(size = 0.5, colour='forestgreen') +
  labs(
    y = 'Number of cells',
    title = 'Number of cells assigned to each barcode',
    subtitle = 'Cell counts computed after projection to 20,000 cells'
  )
plt
```

And for sequencing 10,000 cells?

```{r}
plt <- draw_ordered_abundance_plot(
  count_data = clones_nxclone_proportion,
  y_axis_column = 'projected_10000_confidence_0_8',
  facet_column = 'sample_name'
) + 
  geom_point(size = 0.5, colour='dodgerblue') +
  labs(
    y = 'Number of cells',
    title = 'Number of cells assigned to each barcode',
    subtitle = 'Cell counts computed after projection to 10,000 cells'
  )
plt
```

We can also estimate the number of clones that will contain at least a specified 
minimum number of cells using `count_retained_clones` function used in the previous section.

```{r}
thresholds <- seq(10, 30, 5)

proj_n_clones_retained <- count_retained_clones(
  count_data = clones_nxclone_proportion,
  thresholds = thresholds,
  grouping_col = "sample_name",
  count_column = "projected_10000_confidence_0_8"
)

proj_n_clones_retained
```

As well as the cumulative number of cells for the top say 200 most abundant clones.
This is particularly useful if we want to do differential abundance test for the clones.

```{r}
top_threshold <- 200
top_barcodes <- get_top_barcodes_and_cum_sum(
  count_data = clones_nxclone_proportion,
  count_column = "projected_10000_confidence_0_8",
  grouping_col = "sample_name",
  top_threshold = top_threshold
)

head(top_barcodes)
```

Ah, a table may not be of much use.
Let's draw a line plot to show them instead.

```{r}
ggplot(top_barcodes, aes(x=barcode_rank, y=cum_sum_projected_10000_confidence_0_8, 
                         group=sample_name, colour = sample_name)) +
  geom_line(linewidth=1) +
  theme_bw() +
  scale_y_continuous(breaks = pretty_breaks(n=10), labels = label_comma(accuracy = 1)) +
  scale_x_continuous(breaks = pretty_breaks(n=10)) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    legend.position = "bottom"
  ) +
  labs(
    y = "Number of cells",
    x = "Barcode Ranking",
    title = paste("Projected no. of cells for top", top_threshold, "most abundant clones"),
    colour = "Library"
  )

```







